---
editor_options: 
  markdown: 
    wrap: 72
---

```{r include=FALSE, cache=FALSE}
set.seed(1014)
options(digits = 3)

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  cache = TRUE,
  out.width = "70%",
  fig.align = 'center',
  fig.width = 6,
  fig.asp = 0.618,  # 1 / phi
  fig.show = "hold"
)

options(dplyr.print_min = 6, dplyr.print_max = 6)
```

# Multilevel Models

> "When we go to run our stats and test our hypotheses about the world,
> what if we didn't assume everyone was the same?"
>
> ::: {style="text-align: right"}
> --- Ty Mckinney
> :::

> "The Earth is round (p \< .05)."
>
> ::: {style="text-align: right"}
> --- Jacob Cohen
> :::

First off...

<h4>Advanced R Challenges</h4>

APPLY A "MIXED EFFECTS" MODEL TO YOUR DATASET (if it doesn't make sense
with your data, find a dataset for within and between variables in
"datasets")

## Steps of Multilevel Models {.unnumbered}

## Step 1. Get your data

```{r some data}
require(tidyverse)

df <- read_csv("~/Dropbox/Teaching/MIST130DatAnOpt_22/books_resources/politeness_data.csv")
```

In our study, politeness is treated as a categorical factor with two
levels... a formal register and an informal register. On top of that, we
also had an additional fixed effect, sex, and so our formula looks more
like this:

$$ pitch \sim politeness + sex + ε $$

So far so good. Now things get a little more complicated. Our design was
so that we took multiple measures per subject. That is, each subject
gave multiple polite responses and multiple informal responses. If we go
back to the discussion of the assumptions of the linear model in Chapter
7, we can immediately see that this would violate the independence
assumption: Multiple responses from the same subject cannot be regarded
as independent from each other. Every person has a slightly different
voice pitch, and this is going to be an idiosyncratic factor that
affects all responses from the same subject, thus rendering these
different responses inter-dependent rather than independent.

The way we're going to deal with this situation is to add a random
effect for subject. This allows us to resolve this non-independence by
assuming a different "baseline" pitch value for each subject. So,
subject 1 may have a mean voice pitch of 233 Hz across different
utterances, and subject 2 may have a mean voice pitch of 210 Hz per
subject. Here's a visual depiction of how this looks like:

```{r interdependence}
df %>%
  na.omit() %>%
  group_by(subject,scenario, attitude) %>%
  summarise(mfreq = mean(frequency)) %>%
  ggplot(aes(x = subject, y = mfreq)) + geom_boxplot()
```

We can model these individual differences by assuming different random
intercepts for each subject. That is, each subject is assigned a
different intercept value, and the mixed model estimates these
intercepts for you.

Now you begin to see why the mixed model is called a "mixed" model. The
linear models that we considered so far have been "fixed-effects-only"
models that had one or more fixed effects and a general error term "ε".
With the linear model, we essentially divided the world into things that
we somehow understand or that are somehow systematic (the fixed effects,
or the explanatory variables); and things that we cannot control for or
that we don't understand (ε). But crucially, this latter part, the
unsystematic part of the model, did not have any interesting structure.
We simply had a general across-the-board error term. In the mixed model,
we add one or more random effects to our fixed effects. These random
effects essentially give structure to the error term "ε". In the case of
our model here, we add a random effect for "subject", and this
characterizes idiosyncratic variation that is due to individual
differences. The mixture of fixed and random effects is what makes the
mixed model a mixed model.

Our updated formula looks like this:

$$ pitch \sim politeness + sex + (1|subject) + ε $$

"(1\|subject)" looks a little enigmatic. I'm already using the R-typical
notation format here. What this is saying is "assume an intercept that's
different for each subject" ... and "1" stands for the intercept here.
You can think of this formula as telling your model that it should
expect that there's going to be multiple responses per subject, and
these responses will depend on each subject's baseline level. This
effectively resolves the non-independence that stems from having
multiple responses by the same subject.

Subjects also had to imagine asking a professor for a favor (polite
condition), or asking a peer for a favor (informal condition). An
example of a scenario was: item was an "excusing for coming too late"
scenario, which was similarly divided between polite and informal. In
total, there were 7 such different items.

```{r}
df %>%
  na.omit() %>%
  group_by(subject,scenario, attitude) %>%
  summarise(mfreq = mean(frequency)) %>%
  ggplot(aes(x = scenario, y = mfreq, group = scenario)) + geom_boxplot()




# Challenge 1 -------------------------------------------------------------
 
# What happens if we leave out the "group" argument to aes()?
# Why?


# How do we account for variation in scenario?
```

We do this by adding an additional random effect:

$$ pitch \sim politeness + sex + (1|subject) + (1|item) + ε $$

### Mixed models in R

#### Step 2.

For a start, we need to install the R package lme4 (Bates, Maechler &
Bolker, 2012). open R and type in:

```{r install packages}
# install.packages("lme4")
```

After installation, load the lme4 package into R with the following
command:

```{r load package}
library(lme4)
```

Now, you have the function <code>lmer()</code> available to you, which
is the mixed model equivalent of the function <code>lm()</code> in
tutorial 1. This function is going to construct mixed models for us.

```{r model}
politeness.model = lmer(frequency ~ attitude +
                          (1|subject) + (1|scenario), data=df)


summary(politeness.model)

coef(politeness.model)
```

Let's focus on the output for the random effects first:

```{r random summary}
# Random effects:
#   Groups   Name       Variance Std.Dev.
# scenario (Intercept)  219     14.80   
# subject  (Intercept) 4015     63.36   
# Residual              646     25.42   
# Number of obs: 83, groups:  scenario, 7; subject, 6
```

Have a look at the column standard deviation. This is a measure of how
much variability in the dependent measure there is due to scenarios and
subjects (our two random effects). You can see that scenario ("item")
has much less variability than subject. Based on our boxplots from
above, where we saw more idiosyncratic differences between subjects than
between items, this is to be expected. Then, you see "Residual" which
stands for the variability that's not due to either scenario or subject.
This is our "ε" again, the "random" deviations from the predicted values
that are not due to subjects and items. Here, this reflects the fact
that each and every utterance has some factors that affect pitch that
are outside of the purview of our experiment.

The fixed effects output mirrors the coefficient table that we
considered in tutorial 1 when we talked about the results of our linear
model analysis.

```{r fixed summary}
# Fixed effects:
#              Estimate  Std. Error t value
# (Intercept)  202.588     26.754   7.572
# attitudepol  -19.695      5.585  -3.527
```

The coefficient "attitudepol" is the slope for the categorical effect of
politeness. Minus 19.695 means that to go from "informal" to "polite",
you have to go down-19.695 Hz. In other words: pitch is lower in polite
speech than in informal speech,by about 20 Hz. Then, there's a standard
error associated with this slope, and a tvalue, which is simply the
estimate (20 Hz) divided by the standard error (check this by performing
the calculation by hand).

```{r Challenge 2}
# Challenge 2 -------------------------------------------------------------

# this intercept is especially weird. 
# It’s 202.588 Hz … where does that value come
# from?



politeness.model = lmer(frequency ~ attitude *
                          gender + (1|subject) +
                          (1|scenario), data=df)

summary(politeness.model)
```

Note that compared to our earlier model without the fixed effect gender,
the variation that's associated with the random effect "subject" dropped
considerably. This is because the variation that's due to gender was
confounded with the variation that's due to subject. The model didn't
know about males and females, and so its predictions were relatively
more off, creating relatively larger residuals. Now that we have added
the effect of gender, we have shifted a considerable amount of the
variance that was previously in the random effects component
(differences between male and female individuals) to the fixed effects
component.

```{r random effect summary}
# Random effects:
#   Groups   Name        Variance Std.Dev.
# scenario (Intercept) 219.5    14.81   
# subject  (Intercept) 615.6    24.81   
# Residual             645.9    25.41   
# Number of obs: 83, groups:  scenario, 7; subject, 6

```

We see that males and females differ by about 109 Hz. And the intercept
is now much higher (256.846 Hz), as it now represents the female
category (for the informal condition). The coefficient for the effect of
attitude didn't change much.

```{r model comparison}

# Fixed effects:
#              Estimate Std. Error t value
# (Intercept)  256.846     16.116  15.938
# attitudepol  -19.721      5.584  -3.532
# genderM     -108.516     21.013  -5.164

# Significance

politeness.null = lmer(frequency ~ gender +
                         (1|subject) + (1|scenario), data=df,
                       REML=FALSE)

politeness.model = lmer(frequency ~ attitude +
                          gender + (1|subject) + (1|scenario),
                        data=df, REML=FALSE)

anova(politeness.null,politeness.model)

coef(politeness.model)


politeness.model = lmer(frequency ~ attitude +
                          gender + (1+attitude|subject) +
                          (1+attitude|scenario),
                        data=df,
                        REML=FALSE)



politeness.null = lmer(frequency ~ gender +
                         (1+attitude|subject) + (1+attitude|scenario),
                       data=df, REML=FALSE)

anova(politeness.null,politeness.model)

politeness.model=lmer(frequency~attitude+gender+(1+attitude|scenario), data = df, REML = F)

```

**Challenge 3**

Attempt to apply the mixed model to your own dataset. How many
predictors do you have? Is it appropriate? Why or why not.

**Challenge 4**

What is the intercept of your mixed model? What does it mean for your
data?

**Challenge 5**

Which predictor variables predict your dependent variable? How did you
determine that?

## Solutions

### Challenge 1

What happens if we leave out the "group" argument to aes()?

ggplot() only plots one boxplot.

Why?

The x = scenario only plots x on the x-axis, but geom_boxplot() needs
the additional group argument to plot the scenarios individually.
Otherwise, it groups all of them and plots them together.

How do we account for variation in scenario?

If we have random variation by item, then we need allow each item to
have a random intercept.

### Challenge 2

If you look back at the boxplot that we constructed earlier, you can see
that the value 202.588 Hz seems to fall halfway between males and
females (in the informal condition) -- and this is indeed what this
intercept represents. It's the average of our data for the informal
condition.

### Challenge 3

Attempt to apply the mixed model to your own dataset. How many
predictors do you have? Is it appropriate? Why or why not.

If your dataset has a repeated measures within a person, it is
appropriate. If you do not

### Challenge 4

What is the intercept of your mixed model? What does it mean for your
data?

You should determine if your intercept is meaningful based on your
predictors

### Challenge 5

Which predictor variables predict your dependent variable? How did you
determine that?

Look at the significance tests, such as a p-value.
