---
output:
  pdf_document: default
  html_document: default
---

# Shiny Applications

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will use shiny apps to create dynamic visualizations of data
that can be hosted on websites or from your R console.

**Challenge 1**

Install shiny app:

```{r, eval=FALSE}
install.packages("shiny")
```

Optional installs
```{r, eval=FALSE}

install.packages("shinyjs")
install.packages("shinystan")
install.packages("shinythemes")
```
Open the folder code and open 01-template.R and click run app

Run the lines there to see the structure of a shiny app.

This is the simplest version of a shiny app that can be made.
It creates html to display the text and has no dynamic inputs or
outputs.

Now to 02-hist-app.R


Run the lines there to see the structure of a shiny app.

Now you see inputs and output

**Challenge 2**

Run this shiny app. Screenshot the app.

```{r}


library(shiny)
ui <- fluidPage("hello world") #This is the User Interface (ui),
# it generates html for you so you don't have to do it. Run this line by itself
# and then type ui into your console and hit enter.

#What do you see?

server <- function(input, output) {} #This is where the R code does work.
# Run this line by itself and then hit enter

#What do you see?
shinyApp(ui = ui, server = server) #This starts the app running 
#from you laptop. While it is running, you can see that it has opened
# a port to your local laptop in the console. To stop it, close the
# window or hit the stop sign.

```



```{r}
# library(shiny)

ui <- fluidPage(
  sliderInput(inputId = "num", #Here you can see we set up a slider for the user
    label = "Choose a number", #It has a label, which are instructions
    value = 25, min = 1, max = 100), #It has a default value, a min and a max
  plotOutput("hist"), #And you  have to make sure to plot it
  sliderInput(inputId = "num2", #Here you can see we set up a slider for the user
              label = "Choose a number", #It has a label, which are instructions
              value = 25, min = 1, max = 100), #It has a default value, a min and a max
  plotOutput("hist2") #And you  have to make sure to plot it
)

#This is what ui looks like:
# <div class="container-fluid">
#   <div class="form-group shiny-input-container">
#   <label class="control-label" for="num">Choose a number</label>
#   <input class="js-range-slider" id="num" 
#           data-min="1" data-max="100" data-from="25" 
#           data-step="1" data-grid="true" data-grid-num="9.9" 
#           data-grid-snap="false" data-prettify-separator="," 
#           data-prettify-enabled="true" data-keyboard="true" 
#           data-data-type="number"/>
#   </div>
#   <div id="hist" class="shiny-plot-output" style="width: 100% ; height: 400px"></div>
#   </div>

```

Now for the server function. You use the previously learned R syntax to 
create a function. You give it the arguments input and output.
Then if you have outputs, you need to save the output to "output$name" with the
assignment operator <-

**Challenge 3**

Run this shiny app. Screenshot the app.

```{r}


server <- function(input, output) { #Now we've added input and output to the arguments
  output$hist <- renderPlot({ #output$hist is the saved plot, renderPlot builds it
    hist(rnorm(input$num)) #This creates the histogram with information from input
  })
  output$hist2 <- renderPlot({ #output$hist is the saved plot, renderPlot builds it
    hist(rnorm(input$num2)) #This creates the histogram with information from input  
  }) 
}

shinyApp(ui = ui, server = server)#This creates and runs the app


```

Note: the name in plotOutput() "hist" of the ui function is the same as 
output$hist. These are mapped together to provide the information for the graph.
Also, what you save into output should be something built with a renderName() 
function. You can:

- renderDataTable()
- renderImage()
- renderPlot()
- renderPrint()
- renderTable()
- renderText()
- renderUI()

For outputs you have:

- plotOutput()
- textOutput()
- dataTableOutput()
- imageOutput()
- tableOutput()
- uiOutput()

# 01-kmeans-app

**Challenge 4**

Run this shiny app. Screenshot the app with Petal_Length by Sepal_Width selected.

```{r}

palette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
  "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999"))

library(shiny)

#Let's remember what the dataset iris looks like
View(iris)

ui <- fluidPage(
  headerPanel('Iris k-means clustering'), #This is the title
  sidebarPanel( #This is the sidebar with buttons:
    selectInput('xcol', 'X Variable', names(iris)), #xcol is the value, X Variable the label
    selectInput('ycol', 'Y Variable', names(iris), #ycol is the value, Y Variable the label
      selected = names(iris)[[2]]),
    numericInput('clusters', 'Cluster count', 3,
      min = 1, max = 9)
  ),
  mainPanel(
    plotOutput('plot1')
  )
)


server <- function(input, output) {

  selectedData <- reactive({ #We'll talk about reactives in a bit
    iris[, c(input$xcol, input$ycol)]
  })

  clusters <- reactive({
    kmeans(selectedData(), input$clusters)
  })

  output$plot1 <- renderPlot({ #This creates the plot that you want to see
    #with dynamic changes
    par(mar = c(5.1, 4.1, 0, 1)) #this creates the window position
    plot(selectedData(), #The we plot the clusters from above with selectedData()
         col = clusters()$cluster, #How many clusters
         pch = 20, cex = 3) #Color and size default
    points(clusters()$centers, pch = 4, cex = 4, lwd = 4)
  })

}

shinyApp(ui = ui, server = server)
```


**Challenge 5**

Run your own data with the above code. See if you can replace the numeric variables of the iris dataset with your own. Screenshot the results.
